---
title: "S-PCA"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The octane dataset contains 39 (n)
near-infrared (NIR) absorbance spectra over 226 (p)
different wavelengths. It was 
first discussed in
Esbensen, Sch\"onkopf, and Midtgaard, 1994, 
Multivariate Analysis in Practice, Trondheim: Camo. 
It is known that 
samples 25, 26, and 36:39 contain added alcohol,
and are thus atypical. See
[Hubert, Rousseeuw and Vanden Branden (2005)](http://dx.doi.org/10.1198/004017004000000563)
for more details. 

To compute the S-PCA approximations, first load the necessary functions:
```{r preliminary, message=FALSE, warning=FALSE}
source('S-FPCA-functions.R')
```
A multi-core version of the code is available via the function
`sfpca.par` that distributes the
random starts to different cores, if they are
present. In this example we assume that 
we have 4 cores available. 
```{r prel.par, message=FALSE, warning=FALSE}
library(doParallel)
cl <- makeCluster(4)
registerDoParallel(cl)
```
Now read the data:
```{r read.octane}
x <- read.csv('octane.csv', header=TRUE, dec=',')
x <- x[ ,-(1:2)]
```
Compute an initial location estimate using the L1 estimator, 
implemented in the `pcaPP` package:
```{r initial.centre}
mui <- l1median(X=x, trace=-1)
```
We now compute the 
robustly estimated 2-dimensional
principal subspace, using `Ncand = 1000` random
starts. This takes a long time to compute (about 1hr 
in my computer):
```{r sfpca, cache=TRUE}
q <- 2
Ncand <- 1000
x.s <- sfpca.par(x=x, mu=mui, q=q, Ncand=Ncand, seed=123, init.it=50, max.it=500, 
                 tol=1e-6, trace=FALSE, tuning.rho=3, bb = 0.2426)
```
We now compute the residuals between the
observations and their robust 2-dimensional approximations:
```{r rob.fits}
mu.hat <- x.s$mu
xc <- scale(x, center=x.s$mu, scale=FALSE)
bon <- qr.Q(qr(x.s$b))
x.hat.s <- scale( (xc %*% bon) %*% t( bon ), center=-x.s$mu, scale=FALSE)
re.s <- rowMeans((x-x.hat.s)^2)
```
To determine whether any of these residuals is atypical, 
and considering the natural skewness of squared distances,
we use the adjusted boxplot of 
[Hubert and Vandervieren (2008)](http://dx.doi.org/10.1016/j.csda.2007.11.008)
as implemented in the `robustbase` package:
```{r outliers}
names(re.s) <- 1:nrow(x)
ouS <- as.numeric(names(adjbox(re.s, plot=FALSE)$out))
inliers <- re.s[ouS] < median(re.s)
outs.s <- ouS[!inliers]
outs.true <- c(25, 26, 36:39)
all.equal(outs.s, outs.true)
```
All known outliers are correctly identified. 
We now stop our cluster:
```{r stopcluster}
stopCluster(cl)
```
Look at the original data, the robust predictions (in pink) 
and the detected outliers (in red)
```{r plotdata, echo=FALSE}
matplot(t(x), type='l', col='gray', lty=1, lwd=1.5, ylim=c(0, 0.65), 
        xlab='Frequencies', ylab='Absorbance')
title(main='Octane data')
```
```{r plotpreds, echo=FALSE}
matplot(t(x.hat.s), type='l', col='hotpink', lty=1, lwd=1.5, ylim=c(0, 0.65), 
        xlab='Frequencies', ylab='Absorbance') # , add=TRUE)
title(main='Robust predictions')
```
```{r plotouts, echo=FALSE}
matplot(t(x[ouS,]), type='l', col='red', lty=1, lwd=1.5, ylim=c(0, 0.65), 
        xlab='Frequencies', ylab='Absorbance')
matplot(t(x.hat.s[ouS,]), type='l', col='hotpink', add=TRUE)
title(main='Outliers and their robust predictions')
```

