---
title: "S-PCA"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The octane dataset contains 39 (n)
near-infrared (NIR) absorbance spectra over 226 (p)
different wavelengths. It was 
first discussed in
Esbensen, Sch\"onkopf, and Midtgaard, 1994, 
Multivariate Analysis in Practice, Trondheim: Camo. 
It is known that 
samples 25, 26, and 36:39 contain added alcohol,
and are thus atypical. 

We first load the necessary functions:
```{r preliminary, message=FALSE, warning=FALSE}
source('S-FPCA-functions.R')
```
To utilize multiple cores, we provide the function
`sfpca.par` that distributes the
random starts to different cores, if they are
present. In this example we assume that 
there are 4 cores available. 
```{r prel.par, message=FALSE, warning=FALSE}
library(doParallel)
cl <- makeCluster(4)
registerDoParallel(cl)
```
Now read the data:
```{r read.octane}
x <- read.csv('octane.csv', header=TRUE, dec=',')
x <- x[ ,-(1:2)]
```
Compute an initial centre using the L1 estimator
```{r initial.centre}
mui <- l1median(X=x, trace=-1) # pcaPP::
```
We now compute the projections on the 
robustly estimated 6-dimensional
principal subspace:
```{r sfpca, cache=TRUE}
q <- 3
Ncand <- 100
x.s <- sfpca.par(x=x, mu=mui, q=q, Ncand=Ncand, seed=123, init.it=50, max.it=500, 
                 tol=1e-6, trace=FALSE, tuning.rho=3, bb = 0.2426) 
mu.hat <- x.s$mu
xc <- scale(x, center=x.s$mu, scale=FALSE)
bon <- qr.Q(qr(x.s$b))
x.hat.s <- scale( (xc %*% bon) %*% t( bon ), center=-x.s$mu, scale=FALSE)
re.s <- rowMeans((x-x.hat.s)^2)
#known outliers
outs.true <- c(25, 26, 36:39)
names(re.s) <- 1:nrow(x)
ouS <- as.numeric(names(adjbox(re.s)$out))
inliers <- re.s[ouS] < median(re.s)
(outs.s <- ouS[!inliers])
```
