---
title: "S-estimators for Functional PCA"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## S-estimators for FPCA

This repository contains `R` code to
compute S-estimators for Functional Principal Components Analysis as proposed in 
[Boente and Salibian (2015)](http://dx.doi.org/10.1080/01621459.2014.946991). 

An example applying this method to the octane data set 
can be found [here](octane.md). Code to reproduce the
analysis of the mortality data can be found in
the script `S-FPCA-mortality-example.R`. 
The data is in `FRA_Death_rates_1x2.txt`

The following script illustrates this method when applied to the
well-known lip movement data. 
Click [here](http://www.stats.ox.ac.uk/~silverma/fdacasebook/lipemg.html) 
for more information on these data.
We first load the necessary functions:
```{r preliminary, message=FALSE, warning=FALSE}
source('S-FPCA-functions.R')
```

Read and display the data: the x-axis is time, 
the y-axis is the lip vertical position:
```{r readdata}
# read data
lippo <- read.table('lip-data.txt')
norm <- function(a) sqrt(sum(a^2))
x <- t(lippo)
te <- seq(0,0.69,length=501)
# plot data
matplot(x=te, y=lippo, ylab='', xlab='Time', type='l', col='gray', lty=1, lwd=2)
```

We need the following functions to represent the data on 
a spline basis:
```{r splinefns}
# functions to prepare a spline basis in (0, 0.69) (instead of (0,1))
devolver.base069.sieves <- function(mesh,k)
{    
  aa <- generar.mesh.splines(k)
  knots<-(aa+1)*0.69/2
  base.estim.disc <- cSplineDes(mesh,knots)
  return(base.estim.disc)
}

devolver.base069.sieves.ortonormal <- function(mesh,k)
{
  base.estim.disc <-devolver.base069.sieves(mesh,k)
  base.ortonorm.disc<-qr.Q(qr(base.estim.disc))
  return(base.ortonorm.disc)
}
```
We now compute the coefficients of each observation
on a spline basis of size 20:
```{r splines}
# prepare the spline basis (with 20 functions)
mesh <- te
n <- nrow(x)
dimension.Bspline <- 20 
aa <- generar.mesh.splines(dimension.Bspline )
knots <- (aa+1)*0.69/2
base1 <- devolver.base069.sieves(mesh, dimension.Bspline)
base.estim.Bspline <- devolver.base069.sieves.ortonormal(mesh, dimension.Bspline)
# find the representation of the data on this basis
y <- x %*% base.estim.Bspline
```
We compute an *initial* central function using an L1 estimate: 
```{r initialcentre}
# initial "central" function (L1-estimate)
mui <- l1median(X=y,trace=-1) # pcaPP::
```
We now compute the best 5-dimensional
approximation using an S-estimator. This step may take 
several minutes to complete. 
```{r mainblock, cache=TRUE, warning=FALSE, message=FALSE}
# dimension of the "best" subspace to be estimated 
q <- 5
# number of random starts for the iterative algorithm
Ncand <- 1000 # 1000
y.sfpca <- sfpca(x=y, mu=mui, q=q, Ncand=Ncand, seed=123, init.it=50, max.it=500, 
                 tol=1e-10, trace=FALSE, tuning.rho=3, bb = 0.2426) 
```
Having computed the approximation on the 20-dimensional space of
spline coefficients, we now map the approximations back to the
original space, and plot the corresponding central curve:
```{r mapback}
mu.hat <- as.vector( base.estim.Bspline %*% y.sfpca$mu )
x.hat.ls <- base.estim.Bspline %*% t( y.sfpca$x.ls ) 
yc <- scale(y, center=y.sfpca$mu, scale=FALSE)
bon <- qr.Q(qr(y.sfpca$b))
y.rulo.s <- scale( (yc %*% bon) %*% t( bon ), center=-y.sfpca$mu, scale=FALSE)
x.hat.rulo.s <- base.estim.Bspline %*% t( y.rulo.s ) 
# robust predictions
matplot(x=te, y=x.hat.rulo.s, lty=1, type='l', col='gray', lwd=2, xlab='Time', ylab='')
lines(mu.hat ~ te, lwd=4, col='black')
```

To identify potential outliers, we show the 
squared residuals between the observed curves and their 
predictions (robust (in gray) and classical (in black)):
```{r residuals}
# S- and LS-prediction residuals
re.s <- colMeans((t(x)-x.hat.rulo.s)^2)
re.ls <- colMeans((t(x)-x.hat.ls)^2)

re.s <- re.s * 1e6
re.ls <- re.ls * 1e6

plot(re.s, type='b', pch=19, lwd=3, cex=2.5, col='gray70',ylab="Residual squared norm")
lines(re.ls, type='b', pch=19, lwd=3, cex=2.5, col='black')
abline(h=0.65, lwd=4, lty=2, col='gray30')
```

The following plot shows these potentially outlying curves,
which seem to have their second peak either too late, or 
too early:
```{r outliers}
ous <- (1:n)[re.s > .65]

matplot(x=te, y=t(x), lty=1, type='l', col='gray', xlab='Time', ylab='', lwd=2)
for(i in ous) lines(x=te, y=x[i,], lwd=4, col='gray30')
```






